### 2025-05-16 thesis/thesis.tex
Fix monotonic lang

While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
for NL2SQL systems offer advantages when it comes to human-in-the-loop refinement. Interactive information seeking from the user
has showing to be an effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than
approaching each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require
only small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.

@@

Recent approaches for NL2SQL have increasingly emphasized the generation of multiple candidates and selectiion of the most promising
one amongst them. Candidate selection strategies have shown significant performance improvements on challenging benchmarks. This
approach acknowledges the inherent difficulty of generating perfect SQL queries in a single attempt, even for sophisticated LLMs
and self-correction mechanisms in place.

\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
utilizes execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
characteristics. For candidate selection CHASE utilizes a fine-tuned LLM that can do binary selection of candidates.
\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.

Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which utilizes a multi-generation
ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.

Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generate and training specialized models for candidate
selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
transferability with the robustness of specialized model architectures for candidate ranking and selection.

>>>>

While DIN-SQL and MAGIC focus on automated self-correction in single-turn settings, \cite{CoE-SQL} introduced the concept of
Chain-of-Editions (CoE-SQL), which addresses the unique challenges of multi-turn conversational NLIDBs. Conversational interfaces
for NL2SQL systems enable human-in-the-loop refinement. Interactive information seeking from the user has shown to be an
effective way to drive overall accuracy of the system and improve user satisfaction \citep{NALIR}. Rather than approaching
each query independently, CoE-SQL recognizes that in a conversational context, successive SQL queries usuallly require only
small and incremental modifications of the previous queries. Interactive user input is an effective measure for dealing with
ambiguous natural language queries \citep{CoE-SQL, UnnaturalQueryLanguage, NALIR}.

@@

Contemporary NL2SQL approaches have increasingly emphasized on the generation of query candidates and their selectiion as a
promising architecture. Candidate selection strategies have shown significant performance improvements on challenging benchmarks.
These approach acknowledge the inherent difficulty of generating perfect SQL queries in one attempt / using one generation mechanism,
even with capable LLMs and modern self-correction mechanisms.

\cite{CHASE} introduced CHASE-SQL in \citeyear{CHASE}, a framework that leverages multiple reasoning paths to generate multiple
query candidates. After the initial generation phase CHASE selects the most promising solution to the natural query input.
CHASE-SQL harnesses three different generation strategies: A divide-and-conquer approach which breaks down complex natural
language queries into multiple sub tasks that can be individually tackled, a chain-of-thought based generation approach which
inspects execution plans of SQL queries and a schema-aware generation of synthetic examples that can be used for in-context
learning \citep{CHASE}. These different generation mechanisms produce a set of query candidates that each have different
characteristics. For candidate selection CHASE harnesses a fine-tuned LLM that can do binary selection of candidates.
\citeauthor{CHASE} have demostrated that their query selection approach is more robust than apparent alternatives and yields
state-of-the-art performance with 73\% execution accuracy on \textsc{Bird} and 87.6\% on \textsc{Spider} \citep{CHASE}.

Conceptually similar work has been done by \citeauthor*{XiYan} with \textsc{XiYan-SQL} which is architected as multi-generation
ensemble strategy with better schema representation. \textsc{XiYan-SQL} integrated in-context learning alongside supervised
fine-tuning approaches to generate query candidates \citep{XiYan}. A key contribution of \cite{XiYan} is their M-Schema
representation of database schemas, which improves the models schema awareness and reduces frequent schema linking errors.
\textsc{XiYan-SQL} enhances accuracy by utilizing multiple different strategies that have complementary characteristics
during query generation to enhance the robustness of the overall system. The query generation stage utilizes both a
fine-tuned SQL generation model aswell as ICL strategies to achieve a breadth of candidate coverage. Following to the query
generation stage a self-correction stage (referred to as \textit{refinement} stage by \citeauthor{XiYan}) is utilized
to correct common errors. Lastly a selection model is used to choose the most accurate candidate that was produced
during the generation stage \citep{XiYan}. Through this sophisticated and diverse architecture \textsc{XiYan-SQL} was
able to achieve impressive results across contemporary NL2SQL benchmarks — achieving 89.65\% execution accuracy on
\textsc{Spider} and 73.34\% on \textsc{Bird}, which renders \textsc{XiYan-SQL} state-of-the-art \citep{XiYan}.

Both CHASE and \textsc{XiYan-SQL} show that diversifing candidate generation and training specialized models for candidate
selection yield state-of-the-art execution accuracy which significantly outperforms single-path generation approaches.
The success of these two approaches indicates that for increasingly complex NL2SQL tasks (such as \textsc{Spider2.0}),
the capability to generate multiple valid interpretations of the natural language query is an important stepping stone
to achieving meaningful execution accuracy. Both CHASE and \textsc{XiYan-SQL} rely on specialized candidate selection
models which renders these approaches to combine the strengths of LLMs when it comes to language understanding and
transferability with the robustness of specialized model architectures for candidate ranking and selection.

\vspace{5cm}

% AB HIER WEITER PLS
%CHASE-SQL: Multi-path reasoning and candidate selection \citep{CHASE}
%XiYan-SQL: Multi-generator ensemble framework \citep{XiYan}
%Agent specialization and collaboration patterns
%Voting and selection mechanisms for query ranking

\subsubsection{Retrieval-Augumented Generation}

Integration of retrieval systems with LLMs
DB-GPT: RAG for database interactions \citep{DBGPT}
Schema and documentation retrieval strategies
Impact on handling large/complex database schemas

\subsubsection{Schema Representation}

Chain-of-thought reasoning for complex queries
Schema representation in prompts

% heute goal

\subsubsection{Specialized LLMs and Fine-tuning}

CodeS: Open-source LLMs optimized for SQL \citep{CodeS}
Domain adaptation challenges and solutions \citep{ModelDomainAdaption}
Trade-offs between general-purpose and specialized models
Fine-tuning strategies for improved performance

\subsubsection{Benchmarking}

Contemporary benchmarks: BIRD, Spider 2.0 \citep{BIRD, SPIDER2}
LLM-specific evaluation challenges \citep{LLM-SQL}
Real-world deployment considerations \citep{NL2SQLUnsolved}
Performance metrics beyond accuracy (efficiency, cost, latency)

\subsubsection{Limitations and Challenges}
Hallucination and accuracy concerns
Computational resource requirements
Data privacy and security implications
Ambiguity handling in complex scenarios
Comparison of limitations with PLM approaches

% https://raw.githubusercontent.com/HKUSTDial/NL2SQL_Handbook/refs/heads/main/assets/river.svg
% https://arxiv.org/abs/2109.05093

% CodeS
% GPT-X based
% 
% Prompt engineering and fine tuning

%\subsection{Optimization for LLM-based NL2SQL}
% PICARD, DIN-SQL, MAGIC & CHASE
